Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2013-05-07 CGP Journal/main.tex
Fixed duplicated conclusion
Gave the paper a working title, more fill in

\usepackage{verbatim}
\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{TODO Title}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

@@

To set parameter values, we started by defining a grid of parameter values:
50, 100, 200, 500, 1000, 2000, 5000, 10000 for genome size and 0.05, 0.02, 0.01,
0.005, 0.002, 0.001, 0.0005, 0.0002 for mutation rate.  Note that here a genome
size of 50 means there are 50 nodes, each composed of multiple genes, plus the
required genes to specify output locations.  Also note that mutation rate of 0.05
is used here to mean each gene has a 5\% probability of being mutated in a single
application of mutation.  The grid is formed by trying all possible mutation rates
with all possible genome sizes, resulting in 64 potential configurations for each
of the algorithmic configurations.  Finally note that as \emph{Single} does not
use a mutation rate, only the eight configurations with different genome sizes were
used when \emph{Single} was employed.

@@

TODO Explain reruning.

@@

\section{Analysis of Evolutionary Mechanisms}
\label{sec:analysis}
TODO Explain footprint concept here.

@@

\section{Conclusions}

@@

\section{Conclusion}
The conclusion goes here.~\cite{goldman:2013:cgpwaste}

>>>>

\usepackage{verbatim}
\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

@@

To set parameter values, we started by defining a grid of parameter values:
50, 100, 200, 500, 1000, 2000, 5000, 10000 for genome size and 0.05, 0.02, 0.01,
0.005, 0.002, 0.001, 0.0005, 0.0002 for mutation rate.  Note that here a genome
size of 50 means there are 50 nodes, each composed of multiple genes, plus the
required genes to specify output locations.  Also note that mutation rate of 0.05
is used here to mean each gene has a 5\% probability of being mutated in a single
application of mutation.  These values were chosen as they completely cover the
range of previously used parameter settings, with each value having at least
one paired value for the other parameter which is feasibly effective.
The grid is formed by trying all possible mutation rates
with all possible genome sizes, resulting in 64 potential configurations for each
of the algorithmic configurations.  Finally note that as \emph{Single} does not
use a mutation rate, only the eight configurations with different genome sizes were
used when \emph{Single} was employed.

@@

To ensure termination, all runs were limited to 10,000,000 evaluations, well beyond
the expected time to completion.  As the median was used to compare configurations,
and not the mean, selection should be insensitive to this limit unless a large
portion of runs are failing to optimize.

To prevent undue biasing in the final results, none of these tuning runs were
used for any other data analysis.  This helps reduce selection's effect on
result distribution, similar to how it is common to split training and testing
runs.  In this case, the runs used to set the parameter configuration are considered
the training runs for that configuration.  As a result, our final results were gather
from a completely independent set of runs to test the configurations general characteristics.
Each configuration was run 50 times to ensure statistical power.

@@

\section{Analysis of Evolutionary Mechanisms}
\label{sec:analysis}
Comparing how many evaluations each algorithm takes to solve a set of benchmark
problems only provides us with a limited amount of understanding about how
those algorithms actually function.  We propose here to look for further detail,
both to check on theorized capabilities of each variant and CGP in general, and
to look for previously hidden aspects of CGP.
To do so we have devised a number of novel metrics to examine what makes a
CGP run successful.

As a primary concept to our analysis, we determine the \emph{footprint} of nodes
in the genome.  What we mean by \emph{footprint} is a measure the semantic use
of a node.  For each input that is given to an individual, all of its nodes
will produce a single output.  These output values can be recorded an ordered.
As the domain in use in binary, the outputs can be viewed as a bit string, with
each bit the value that a node outputs when an individual is presented with a
specific input.  Similar recording can be done for other domains, but binary
allows for by far the simplest encoding.

The \emph{footprint} of a node is a complete description of its functionality.
Any two nodes with identical \emph{footprint} values have identical behavior,
even if their method for encoding that behavior is drastically different.  From
this perspective, we can view CGP evolution as attempting to evolve a node who's
\emph{footprint} matches the desired output \emph{footprint}.  Doing so allows
us to examine how the evolutionary operators change the behavior of nodes,
not just their gene values.

Utilizing \emph{footprint} analysis, we can extract the semantically useful
portions of an individual, a subset of the active nodes.  Consider two active
nodes with the same \emph{footprint}.  As CGP is a graph representation, only
one of these nodes is actually necessary.  Given one in the genome, any node that reads
from the other can have its connection changed without changing its behavior.

TODO Explain simplify algorithm

@@

@@

\section{Conclusion}
The conclusion goes here.
